---
title: "Modeling plant height"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling plant height}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use a double exponential model and 
plant height data derived from UAV imagery to estimate two key parameters:

* t2: the number of days it takes to reach the maximum plant height.
* f(t2): the maximum plant height.

Before fitting any model for plant height, we will first reference the previous 
vignette  [Modeling Canopy](https://apariciojohan.github.io/flexFitR/articles/canopy-model.html)
to estimate initial values for emergence, which will be used in the double-exponential model.

The data in this vignette are part of Professor Jeff Endelman's breeding program,
specifically from a partially replicated experiment. The UAV images were collected
in 2022 and processed in 2024.

## Loading libraries

```{r, warning=FALSE, message=FALSE }
library(flexFitR)
library(dplyr)
library(kableExtra)
library(ggpubr)
library(purrr)
```

## 1. Exploring data

We start by exploring the data using the explorer function. This provides basic statistical summaries and visualizations, which help in understanding the temporal evolution of the variables for each plot and their correlations.

```{r}
data(dt_potato_22)
results <- explorer(dt_potato_22, x = DAP, y = c(PH, Canopy), id = Plot)
```

```{r}
names(results)
```

```{r, fig.width= 8, fig.height=6, fig.alt="plot corr"}
p1 <- plot(results, type = "evolution", return_gg = TRUE, add_avg = TRUE)
p2 <- plot(results, type = "x_by_var", return_gg = TRUE)
ggarrange(p1, p2, nrow = 2)
```
```{r}
kable(mutate_if(filter(results$summ_vars, var == "PH"), is.numeric, round, 2))
```


## 2. Estimating days to emergence

As previously shown in the canopy model vignette, we use a piece-wise regression function with three parameters: t1, t2, and k. This can be used to estimate the day of plant emergence (t1), which will later be used as input for the double-exponential model for plant height.

\begin{equation}
f(t; t_1, t_2, k) =
  \begin{cases}
0 & \text{if } t < t_1 \\
\dfrac{k}{t_2 - t_1} \cdot (t - t_1) & \text{if } t_1 \leq t \leq t_2 \\
k & \text{if } t > t_2
\end{cases}
\end{equation}

```{r, echo = FALSE, fig.width= 8, fig.alt="plot fn"}
plot_fn(
  fn = "fn_piwise",
  params = c(t1 = 34.9, t2 = 61.8, k = 100),
  interval = c(0, 108),
  color = "black",
  base_size = 15
)
```


## 2.1. Fitting models for canopy

The parameters find here will serve as fixed parameters in the plant height model.

```{r}
fixed_params <- data.frame(uid = c(195, 40), k = c(100, 100))
```

```{r, warning=FALSE, message=FALSE}
mod_1 <- dt_potato_22 |>
  modeler(
    x = DAP,
    y = Canopy,
    grp = Plot,
    fn = "fn_piwise",
    parameters = c(t1 = 45, t2 = 80, k = 0.9),
    fixed_params = fixed_params,
    subset = c(195, 40),
    options = list(add_zero = TRUE, max_as_last = TRUE)
  )
```

```{r, fig.width= 8, fig.height=4, fig.alt="plot fit"}
plot(mod_1, id = c(195, 40))
kable(mod_1$param)
```

## 3. Expectation function for plant height 

Once the data are explored, we define the expectation function for plant height, 
which in this case is a double-exponential function with four parameters: 
t1, t2, $\alpha$, and $\beta$. This function models the growth dynamics of plant height over time:

`fn_exp2_exp()`

\begin{equation}
f(t; t_1, t_2, \alpha, \beta) =
\begin{cases}
0 & \text{if } t < t_1 \\
e^{\alpha \cdot (t - t_1)^2} - 1 & \text{if } t_1 \leq t \leq t_2 \\
\left(e^{\alpha \cdot (t_2 - t_1)^2} - 1\right) \cdot e^{\beta \cdot (t - t_2)} & \text{if } t > t_2
\end{cases}
\end{equation}

```{r, echo = FALSE, fig.width= 8, fig.alt="plot fn"}
plot_fn(
  fn = "fn_exp2_exp",
  params = c(t1 = 35, t2 = 55, alpha = 1 / 600, beta = -1 / 30),
  interval = c(0, 108),
  color = "black",
  base_size = 15
)
```

## 4. Fixing parameters and providing initial values

Before fitting the plant height model, we take the t1 values from the canopy model
and use them as fixed parameters in the plant height model. This ensures consistency between the two models.

```{r}
fixed_params <- mod_1 |>
  pluck("param") |>
  select(uid, t1)
kable(fixed_params)
```

Additionally, we can specify initial values for the parameters of each plot to improve the model's convergence.

```{r}
initials <- mod_1 |>
  pluck("param") |>
  select(uid, t1, t2) |>
  mutate(alpha = 1 / 600, beta = -1 / 30)
kable(initials)
```

## 5. Fitting models for plant height

To fit the model, we use the modeler function. Here:

* x specifies the days after planting (DAP),
* y is the plant height variable to be modeled,
* grp is used for grouping, allowing analysis by plot.

In this example, although there are 196 plots, we will fit the model for plots 195 and 40 as a subset.
The `fn_exp2_exp` function is defined, and we set initial values for the parameters.

```{r}
mod_2 <- dt_potato_22 |>
  modeler(
    x = DAP,
    y = PH,
    grp = Plot,
    fn = "fn_exp2_exp",
    parameters = initials,
    fixed_params = fixed_params,
    subset = c(195, 40),
    options = list(add_zero = TRUE)
  )
```

After fitting, we can inspect the model summary and visualize the fit using the plot function:

```{r, fig.width= 8, fig.height=4, fig.alt="plot fit 2"}
plot(mod_2, id = c(195, 40))
kable(mod_2$param)
```


## 6. Extracting model coefficients and uncertainty measures

Once the model is fitted, we can extract key statistical information, such as the
estimated coefficients, standard errors, confidence intervals, and the 
variance-covariance matrix for each plot. This helps evaluate the reliability and
uncertainty of the parameter estimates.

The functions `coef`, `confint`, and `vcov` are used as follows:

* **coef**: Extracts the estimated coefficients for each group.
* **confint**: Provides the confidence intervals for the parameter estimates.
* **vcov**: Returns the variance-covariance matrix, which can be used to understand 
the relationships between the estimates and their variability.

```{r}
coef(mod_2)
```
```{r}
confint(mod_2)
```

```{r}
vcov(mod_2)
```

## 7. Predicting maximun plant height

Once we have estimated t2, which indicates the number of days it takes to reach
maximum plant height, we can use the predict function to calculate the expected 
maximum height at that specific time point.

```{r}
# Maximum Plant Height
predict(mod_2, x = 64.5550589254, id = 40)
predict(mod_2, x = 62.0000000000, id = 195)
```
In this example, we predict the maximum plant height for plot 40 at approximately 64.56 DAP
and for plot 195 at 62.00 DAP.

## 8. Modelling all plots using parallel processing

Finally, we can scale up this method to fit models for all 196 plots,
using parallel processing to accelerate the computation. By setting
`parallel = TRUE` in the options argument and specifying the number of cores 
with `parallel::detectCores()`, the process becomes much more efficient.

```{r, eval= FALSE}
mod <- dt_potato_22 |>
  modeler(
    x = DAP,
    y = PH,
    grp = Plot,
    fn = "fn_exp2_exp",
    parameters = initials,
    fixed_params = fixed_params,
    subset = c(195, 40),
    options = list(
      add_zero = TRUE,
      max_as_last = TRUE,
      progress = TRUE,
      parallel = TRUE,
      workers = parallel::detectCores()
    )
  )
```

<hr>
