---
title: "Modelling Plant Height"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modelling Plant Height}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Loading libraries

```{r, warning=FALSE, message=FALSE }
library(exploreHTP)
library(dplyr)
library(kableExtra)
library(ggpubr)
library(purrr)
```

## Reading data `read_HTP()`

```{r}
data(dt_chips)
results <- read_HTP(dt_chips, x = DAP, y = c(PH, Canopy), id = Plot)
```

```{r}
names(results)
```

## Data exploration

```{r, fig.width= 8, fig.height=6}
p1 <- plot(results, type = "evolution", return_gg = TRUE)
p2 <- plot(results, type = "time_by_trait", return_gg = TRUE)
ggarrange(p1, p2, nrow = 2)
```

## Estimating days to emergence

\begin{equation}
f(t; t_1, t_2, k) =
  \begin{cases}
0 & \text{if } t < t_1 \\
\dfrac{k}{t_2 - t_1} \cdot (t - t_1) & \text{if } t_1 \leq t \leq t_2 \\
k & \text{if } t > t_2
\end{cases}
\end{equation}

```{r, echo = FALSE, fig.width= 8}
plot_fn(
  fn = "fn_piwise",
  params = c(t1 = 34.9, t2 = 61.8, k = 100),
  interval = c(0, 108),
  color = "black",
  base_size = 15
)
```


## Fitting models for canopy

```{r}
fixed_params <- results |>
  pluck("dt_long") |>
  filter(var %in% "Canopy" & uid %in% c(195, 40)) |>
  group_by(uid) |>
  summarise(k = max(y), .groups = "drop")
```

```{r, warning=FALSE, message=FALSE}
mod_1 <- dt_chips |>
  modeler_HTP(
    x = DAP,
    y = Canopy,
    by = Plot,
    id = c(195, 40),
    parameters = c(t1 = 45, t2 = 80, k = 0.9),
    fn = "fn_piwise",
    fixed_params = fixed_params,
    add_zero = TRUE,
    max_as_last = TRUE
  )
```

```{r, fig.width= 8, fig.height=4}
plot(mod_1, id = c(195, 40))
kable(mod_1$param)
```

## Target function for plant height 

`fn_exp2_exp()`

\begin{equation}
f(t; t_1, t_2, \alpha, \beta) =
\begin{cases}
0 & \text{if } t < t_1 \\
e^{\alpha \cdot (t - t_1)^2} - 1 & \text{if } t_1 \leq t \leq t_2 \\
\left(e^{\alpha \cdot (t_2 - t_1)^2} - 1\right) \cdot e^{\beta \cdot (t - t_2)} & \text{if } t > t_2
\end{cases}
\end{equation}

```{r, echo = FALSE, fig.width= 8}
plot_fn(
  fn = "fn_exp2_exp",
  params = c(t1 = 35, t2 = 55, alpha = 1 / 600, beta = -1 / 30),
  interval = c(0, 108),
  color = "black",
  base_size = 15
)
```

## Fixing parameters and providing initial values

```{r}
fixed_params <- mod_1 |>
  pluck("param") |>
  select(uid, t1)
kable(fixed_params)
```

```{r}
initials <- mod_1 |>
  pluck("param") |>
  select(uid, t1, t2) |>
  mutate(alpha = 1 / 600, beta = -1 / 30)
kable(initials)
```


## Fitting models for plant height

Using t1 from the previous model as a fixed parameter and using t2 as initial parameter per plot.


```{r}
mod_2 <- dt_chips |>
  modeler_HTP(
    x = DAP,
    y = PH,
    by = Plot,
    id = c(195, 40),
    fn = "fn_exp2_exp",
    initial_vals = initials,
    fixed_params = fixed_params,
    add_zero = TRUE
  )
```


```{r, fig.width= 8, fig.height=4}
plot(mod_2, id = c(195, 40))
kable(mod_2$param)
```

<hr>

## Code

```{r, eval = FALSE}
data(dt_chips)
results <- read_HTP(dt_chips, x = DAP, y = c(PH, Canopy), id = Plot)
fixed_params <- results |>
  pluck("dt_long") |>
  filter(var %in% "Canopy" & uid %in% c(195, 40)) |>
  group_by(uid) |>
  summarise(k = max(y), .groups = "drop")
mod_1 <- dt_chips |>
  modeler_HTP(
    x = DAP,
    y = Canopy,
    by = Plot,
    id = c(195, 40),
    parameters = c(t1 = 45, t2 = 80, k = 0.9),
    fn = "fn_piwise",
    fixed_params = fixed_params,
    add_zero = TRUE,
    max_as_last = TRUE
  )
fixed_params <- mod_1 |>
  pluck("param") |>
  select(uid, t1)
initials <- mod_1 |>
  pluck("param") |>
  select(uid, t1, t2) |>
  mutate(alpha = 1 / 600, beta = -1 / 30)
mod_2 <- dt_chips |>
  modeler_HTP(
    x = DAP,
    y = PH,
    by = Plot,
    id = c(195, 40),
    fn = "fn_exp2_exp",
    initial_vals = initials,
    fixed_params = fixed_params,
    add_zero = TRUE
  )
```
