---
title: "Modeling vegetation index (VI)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling vegetation index (VI)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use the flexFitR package to model a vegetation 
index derived from UAV imagery. We will focus on the Green Leaf Index (GLI) to 
estimate three key events in plant development:

* t1: The day of plant emergence.
* t2: The number of days it takes to reach canopy closure.
* t3: The number of days it takes to onset of senescence.

The data presented here are part of Professor Jeff Endelman's breeding program,
specifically for a partially replicated experiment. The UAV images were collected
in 2020 and processed in 2024.

## Loading libraries

```{r, warning=FALSE, message=FALSE }
library(flexFitR)
library(dplyr)
library(kableExtra)
library(ggpubr)
library(purrr)
```

## 1. Exploring data

We begin by exploring the data using the `explorer` function. This function provides basic statistical summaries and visualizations, helping us understand the temporal evolution of each plot and the correlations among the variables.

```{r}
data(dt_potato)
explorer <- explorer(dt_potato, x = DAP, y = c(Canopy, GLI_2), id = Plot)
```

```{r}
names(explorer)
```

```{r, fig.width= 8, fig.height=6, fig.alt="plot corr"}
p1 <- plot(explorer, type = "evolution", return_gg = TRUE, add_avg = TRUE)
p2 <- plot(explorer, type = "x_by_var", return_gg = TRUE)
ggarrange(p1, p2, nrow = 2)
```

```{r, echo=FALSE}
kable(mutate_if(filter(explorer$summ_vars, var == "GLI_2"), is.numeric, round, 2))
```

## 2. Expectation function


After exploring the data, we define the expectation function. In this case, we 
utilize a linear-plateau-linear function characterized by five parameters: 
t1, t2, t3, k, and $\beta$. The function can be expressed mathematically as follows:

`fn_lin_pl_lin()`

\begin{equation}
f(t; t_1, t_2, t_3, k, \beta) =
\begin{cases}
0 & \text{if } t < t_1 \\
\dfrac{k}{t_2 - t_1} \cdot (t - t_1) & \text{if } t_1 \leq t \leq t_2 \\
k & \text{if } t_2 \leq t \leq t_3 \\
k + \beta \cdot (t - t_3) & \text{if } t > t_3
\end{cases}
\end{equation}

```{r, echo = FALSE, fig.width= 8,fig.width= 5, fig.alt="plot fn"}
plot_fn(
  fn = "fn_lin_pl_lin",
  params = c(t1 = 38.7, t2 = 62, t3 = 90, k = 0.32, beta = -0.01),
  interval = c(0, 108),
  color = "black",
  base_size = 15
)
```

## 3. Fitting Models

To fit the model, we use the modeler function. Here:

* x specifies the days after planting (DAP),
* y the GLI_2 variable that we aim to model.
* grp Allows for group analysis, enabling us to analyze multiple plots.

In this vignette, we will fit the model for plots 195 and 40 as a subset, out of
a total of 196 plots. We will define the linear-plateau-linear function 
`fn_lin_pl_lin` and set initial values for the parameters. Additionally,
we can include a zero value in the time series if needed.

```{r, warning=FALSE, message=FALSE}
mod_1 <- dt_potato |>
  modeler(
    x = DAP,
    y = GLI_2,
    grp = Plot,
    fn = "fn_lin_pl_lin",
    parameters = c(t1 = 38.7, t2 = 62, t3 = 90, k = 0.32, beta = -0.01),
    subset = c(195, 40),
    options = list(add_zero = TRUE)
  )
```

After fitting, we can inspect the model summary and visualize the fit using the plot function:

```{r, fig.width= 8, fig.height=4, fig.alt="plot fit 1"}
plot(mod_1, id = c(195, 40))
kable(mod_1$param)
```

## 3.1. Extracting model coefficients and uncertainty measures

Once the model is fitted, we can extract essential statistical information,
including coefficients, standard errors, confidence intervals, and the 
variance-covariance matrix for each plot. These metrics help us evaluate the 
reliability of the parameter estimates and assess the uncertainty around them.

The functions `coef`, `confint`, and `vcov` are used as follows:

* **coef**: Extracts the estimated coefficients for each group.
* **confint**: Provides the confidence intervals for the parameter estimates.
* **vcov**: Returns the variance-covariance matrix, which can be used to understand 
the relationships between the estimates and their variability.

```{r}
coef(mod_1)
```
```{r}
confint(mod_1)
```

```{r}
vcov(mod_1)
```

## 4. Adding constrains to the optimization

To impose constraints on the optimization process, we can modify the expectation 
function. In this particular case, if we want to ensure that t3 is greater than
or equal to t2, we can reformulate the function, introducing dt as the difference
between t3 and t2:

\begin{equation}
f(t; t_1, t_2, dt, k, \beta) =
\begin{cases}
0 & \text{if } t < t_1 \\
\dfrac{k}{t_2 - t_1} \cdot (t - t_1) & \text{if } t_1 \leq t \leq t_2 \\
k & \text{if } t_2 \leq t \leq (t_2 + dt) \\
k + \beta \cdot (t - (t_2 + dt)) & \text{if } t > (t_2 + dt)
\end{cases}
\end{equation}

We can then add constraints in the modeler function using the lower and upper 
arguments, where the coefficient dt can vary from 0 to infinity, and the slope
($\beta$) must be non-positive.

```{r}
mod_2 <- dt_potato |>
  modeler(
    x = DAP,
    y = GLI_2,
    grp = Plot,
    fn = "fn_lin_pl_lin2",
    parameters = c(t1 = 38.7, t2 = 62, dt = 28, k = 0.32, beta = -0.01),
    lower = c(-Inf, -Inf, 0, -Inf, -Inf),
    upper = c(Inf, Inf, Inf, Inf, 0),
    method = c("nlminb", "L-BFGS-B"),
    subset = c(195, 40),
    options = list(add_zero = TRUE)
  )
```

```{r, fig.width= 8, fig.height=4 , fig.alt="plot fit 4"}
plot(mod_2, id = c(195, 40))
kable(mod_2$param)
```

```{r}
coef(mod_1)
```
```{r}
confint(mod_1)
```

```{r}
vcov(mod_1)
```

<hr>
