---
title: "Model selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Choosing a different model for every plot

Canopy development can vary considerably among plots. Some follow an approximately linear pattern, whereas others show a more sigmoidal response. As a result, a single functional form may describe some plots well but provide a poor fit for others.

The `model_selection()` function allows a different model to be selected for each plot. 
Candidate models are fitted to the same data, and the best-fitting model is selected according to a specified criterion.
The results are returned as a single `modeler` object containing the selected fit for each plot.

We use canopy cover data derived from UAV imagery collected by the University of Wisconsin–Madison Potato Breeding Program. The dataset includes eight flights conducted between 0 and 100 days after planting (DAP).

```{r, warning=FALSE, message=FALSE}
library(flexFitR)
library(dplyr)
library(kableExtra)
library(ggplot2)
```

## 1. Exploring data

```{r}
data(dt_potato)
explorer <- explorer(dt_potato, x = DAP, y = Canopy, id = Plot)
```

```{r, fig.width = 8, fig.height = 4, fig.alt = "Canopy cover against days after planting for every plot, with the trial average overlaid"}
plot(explorer, type = "evolution", add_avg = TRUE)
```

Canopy cover increases rapidly between approximately 30 and 60 DAP, before approaching a plateau near 100%. The individual trajectories show some variation in the timing and shape of this increase.

## 2. Candidate models

We consider three models: linear-plateau, logistic, and quadratic-plateau. Each describes an increase in canopy cover toward a maximum value, but they differ in how they represent the growth phase and the transition to the plateau.

**Linear-plateau** — This three-parameter model describes a linear increase in canopy cover from ($t_1$) to ($t_2$), followed by a constant plateau at ($k$):

\begin{equation}
f(t; t_1, t_2, k) =
\begin{cases}
0 & \text{if } t < t_1 \\
\dfrac{k}{t_2 - t_1} \cdot (t - t_1) & \text{if } t_1 \leq t \leq t_2 \\
k & \text{if } t > t_2
\end{cases}
\end{equation}

**Logistic** — This three-parameter model describes sigmoidal growth, with a gradual increase followed by a slowdown as canopy cover approaches the asymptote ($k$). The parameter ($t_0$) represents the inflection point, and ($a$) controls the growth rate:

\begin{equation}
f(t; a, t_0, k) = \dfrac{k}{1 + e^{-a (t - t_0)}}
\end{equation}

**Quadratic-plateau** — This four-parameter model describes a quadratic increase in canopy cover from ($t_1$) to ($t_2$), followed by a constant plateau at ($k$). The parameter ($b$) represents the initial slope of the growth phase.:

\begin{equation}
f(t; t_1, t_2, b, k) =
\begin{cases}
0 & \text{if } t < t_1 \\
b (t - t_1) + \dfrac{k - b (t_2 - t_1)}{(t_2 - t_1)^2} (t - t_1)^2 & \text{if } t_1 \leq t \leq t_2 \\
k & \text{if } t > t_2
\end{cases}
\end{equation}


We use `plot_fn()` to visualize each model with the starting parameter values that will be used for fitting:

```{r, fig.width = 8, fig.height = 3, fig.alt = "The linear-plateau function drawn at its starting values"}
plot_fn("fn_lin_plat", params = c(t1 = 45, t2 = 80, k = 90), interval = c(0, 100))
```

```{r, fig.width = 8, fig.height = 3, fig.alt = "The logistic function drawn at its starting values"}
plot_fn("fn_logistic", params = c(a = 0.199, t0 = 47.7, k = 100), interval = c(0, 100))
```

```{r, fig.width = 8, fig.height = 3, fig.alt = "The quadratic-plateau function drawn at its starting values"}
plot_fn("fn_quad_plat", params = c(t1 = 45, t2 = 80, b = 5, k = 100), interval = c(0, 100))
```

## 3. Fitting the candidates

For this example, we fit the three models to a subset of six plots. The `subset` argument can be omitted to fit the models to the entire trial.

```{r, warning = FALSE, message = FALSE}
ids <- c(1, 7, 15, 66, 84, 89)
mod_1 <- dt_potato |>
  modeler(
    x = DAP, 
    y = Canopy,
    grp = Plot,
    fn = "fn_lin_plat",
    parameters = c(t1 = 45, t2 = 80, k = 90), 
    subset = ids
  )
mod_2 <- dt_potato |>
  modeler(
    x = DAP,
    y = Canopy,
    grp = Plot, 
    fn = "fn_logistic",
    parameters = c(a = 0.199, t0 = 47.7, k = 100), 
    subset = ids
  )
mod_3 <- dt_potato |>
  modeler(
    x = DAP, 
    y = Canopy,
    grp = Plot,
    fn = "fn_quad_plat",
    parameters = c(t1 = 45, t2 = 80, b = 1, k = 100),
    subset = ids
  )
```

## 4. Comparing by hand with `performance()`

We use `performance()` to compare the candidate models using AIC, AICc, and BIC in this example.

```{r}
comparison <- performance(mod_1, mod_2, mod_3, metrics = c("AIC", "AICc", "BIC"))

comparison |>
  filter(uid %in% c(1, 7)) |>
  kable(caption = "Information criteria for the first two plots")
```

The quadratic-plateau model has four curve parameters, compared with three for the linear-plateau and logistic models.
Including the residual variance, the corresponding parameter counts used for AIC and AICc are five and four, respectively. 
With only eight observations per plot, the small-sample correction in AICc is substantial: it adds 30 to the AIC of the quadratic-plateau model and 13.3 to the other two models. As a result, AICc may favor a simpler model even when the more complex model has a lower AIC.

The radar plot rescales every metric within a plot, so the outer edge is always
the best model:

```{r, fig.width = 8, fig.height = 4, fig.alt = "Radar plot comparing the three candidate models across information criteria for two plots"}
plot(comparison, id = c(1, 7), type = 1)
```

Comparing models individually becomes impractical when the number of plots is large. 
The `model_selection()` function automates this process.


## 5. Automatic selection

The fitted models are supplied along with the selection criterion, which in this example is "AICc".

```{r}
best <- model_selection(mod_1, mod_2, mod_3, metrics = "AICc")
best
```

The function returns a single `modeler` object containing the selected fit for each plot. `attr(best, "selection")` records which  model was selected for each plot, along with its selection score.

```{r}
attr(best, "selection") |>
  kable(caption = "Best model per plot, with its score")
```

The `score` column summarizes the model's performance across the selected criteria. Each criterion is rescaled from 0.1 to 1, with higher values indicating better performance, and the scores are averaged when multiple criteria are used. With a single criterion, the selected model receives a score of 1, so the score is mainly useful when combining several criteria. The `n_metrics` column indicates how many criteria contributed to the score, which is relevant when a criterion cannot be calculated for a particular plot.

Use `return_table` if you want the table rather than the refitted object:

```{r}
model_selection(mod_1, mod_2, mod_3, metrics = "AICc", return_table = TRUE) |>
  count(model, name = "groups") |>
  kable(caption = "How often each candidate won")
```

The linear-plateau model was selected for four of the six plots, while the remaining plots favored other models. This illustrates that the preferred model can vary among plots within the same trial, and that fitting a single functional form to all plots may not adequately describe their individual growth trajectories.

The complete comparison across candidate models is also stored in the performance attribute of the returned object and can be accessed or plotted directly.

```{r, fig.width = 8, fig.height = 4, fig.alt = "plot"}
plot(attr(best, "performance"), type = 3, id = ids)
```

## 6. The selected object behaves like any other fit

Because `model_selection()` returns a `modeler` object, the usual methods can be applied directly to the selected fits.

```{r}
predict(best, x = 60) |>
  head(6) |>
  kable(caption = "Predicted canopy at 60 DAP, each plot from its own model")
```

```{r, fig.width = 9, fig.height = 4, fig.alt = "Observed canopy and fitted curves for 3 plots, each fitted with its selected model"}
plot(best, id = c(1, 7, 66)) + theme(legend.position = "top")
```

## 7. Which criterion should you rank on?

`AICc` is used as the default selection criterion. In this example, each plot contains only eight observations, so the small-sample correction can have a substantial effect on model ranking. To illustrate this, we compare the models selected using AIC and AICc.

```{r, warning = FALSE}
by_aic <- model_selection(mod_1, mod_2, mod_3, metrics = "AIC", return_table = TRUE)
by_aicc <- model_selection(mod_1, mod_2, mod_3, metrics = "AICc", return_table = TRUE)

inner_join(by_aic, by_aicc, by = "uid", suffix = c("_aic", "_aicc")) |>
  count(model_aic, model_aicc, name = "plots") |>
  kable(caption = "AIC against AICc: where the two criteria disagree")
```

In this example, the disagreements occur when AIC selects the more flexible quadratic-plateau model, whereas AICc selects a simpler alternative. With only eight observations per plot, the additional small-sample penalty in AICc has a strong influence on model selection. For this reason, AICc is more appropriate than AIC for these data.

Multiple selection criteria can also be used simultaneously. In that case, each metric is rescaled within a plot and the resulting values are averaged to obtain an overall score for each candidate model.

```{r, warning = FALSE}
model_selection(mod_1, mod_2, mod_3, metrics = c("AICc", "BIC")) |>
  attr("selection") |>
  count(model, name = "plots") |>
  kable(caption = "Ranking on AICc and BIC together")
```

The option `metrics = "all"` should be used with care because several of the available metrics contain overlapping information. For example, SSE, MSE, RMSE, and \(R^2\) produce equivalent rankings within a plot, while AIC, AICc, and BIC are all derived from the likelihood with different penalties for model complexity. Including all metrics therefore gives greater influence to aspects of model performance that are represented by several related criteria. When combining metrics, it is preferable to select a subset that reflects the aspects of model performance relevant to the analysis.

## 8. When a plot cannot be compared

For a given plot, model selection requires at least one criterion that can be evaluated across the candidate models. If a metric is unavailable for one of the candidates, that metric is excluded from the comparison for that plot. If no valid criteria remain, the plot is omitted from the selection results and a warning is returned.

```{r}
dropped <- setdiff(unique(mod_1$param$uid), attr(best, "selection")$uid)
dropped

performance(mod_1, mod_2, mod_3, metrics = c("logLik", "AIC", "AICc", "SSE")) |>
  filter(uid %in% dropped) |>
  kable(caption = "The candidate fits for the dropped plot")
```

For this plot, the quadratic-plateau model reproduces all eight observations exactly, resulting in an SSE of zero. Under the Gaussian likelihood used here, this causes the log-likelihood to approach \(+\infty\), while AIC, AICc, and BIC approach \(-\infty\). These values should not be interpreted as evidence of an exceptionally good model; rather, they indicate that the model has interpolated the observed data exactly.

```{r}
model_selection(mod_1, mod_2, mod_3, metrics = "RMSE", return_table = TRUE) |>
  filter(uid %in% dropped) |>
  kable(caption = "The same plot, ranked on RMSE")
```

Residual-based metrics remain finite for this plot, so the candidates can still be ranked using a criterion such as RMSE. However, because RMSE does not penalize model complexity, the exactly interpolating quadratic-plateau model will be favored. This example therefore also illustrates why the choice of selection criterion should consider both goodness of fit and model complexity.

## Summary

* Fit each candidate model first, then use `model_selection()` to select the preferred model separately for each plot.
* `model_selection()` returns a `modeler` object, so methods such as `predict()`, `coef()`, `vcov()`, `confint()`, `inverse_predict()`, `augment()`, `compute_tangent()`, and `plot()` can be applied directly to the selected fits.
* The `selection` attribute records the model selected for each plot, while the `performance` attribute retains the complete comparison among candidate models.
* When the number of observations per plot is small relative to the number of estimated parameters, AICc is generally more appropriate than AIC. When combining several criteria, select metrics carefully because some provide overlapping information.
* If a criterion cannot be evaluated across the candidate models for a plot, it is excluded from that comparison. If no valid criteria remain, the plot is omitted from the selection results and a warning is returned.
